---
title: "Homework 02: Explore Gapminder and use dplyr"
output: 
    github_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the gapminder and tidyverse library
```{r}
library(gapminder)
library(tidyverse)
```

### Smell test the data
Object gapminder is a list. It can be determined by "typeof()" command:
```{r}
typeof(gapminder)
```

Object gapminder is a class of tbl_df. 
> tbl_df object is a data frame providing a nicer printing method, useful when working with large data sets.
```{r}
class(gapminder)
```

It has 6 variables/columns and 1704 rows/observations.
```{r}
ncol(gapminder)
nrow(gapminder)
```

Another way to get the size is to use dim(x)
```{r}
dim(gapminder)
```

"sapply(list, function)" with function = typeof can be used to display the data type of each variable
```{r}
sapply(gapminder, typeof)
```

### Explore individual variables
#### Explore variable "lifeExp"
Getting range of lifeExp seems appropriate. To do so:
```{r}
gapminder %>% 
  select(lifeExp) %>% 
  range()
```

Use "summary()" to get the mean, min, max and other statistics
```{r}
summary(gapminder$lifeExp)
```

Graphical statistics can be used to get the distribution and spread of lifeExp
```{r}
ggplot(gapminder, aes(lifeExp)) + 
  geom_histogram(aes(y=..density..), color = 'blue', fill = 'orange', bins = 50) + 
  geom_density()
```

#### Explore variable gdpPercap
summary of gdpPercap
```{r}
summary(gapminder$gdpPercap)
```

Distribution of gdpPercap
```{r}
ggplot(gapminder, aes(gdpPercap)) + 
  geom_histogram(aes(y=..density..), color = 'black', fill = 'blue', bins = 50) + 
  geom_density()
```

#### Explore variable country
For categorical variable like country, "unique()" can be used to determine what countries are studied. We can see 142 different countries are in gapminder
```{r}
unique(gapminder$country)
```

Statistics about country is not very useful. However, we could filter by country to study the other variables against the same country. 
```{r}
gapminder %>% 
  filter(country == 'Germany') %>% 
  ggplot(aes(year, lifeExp)) + 
  geom_line() + 
  geom_point()
```

### Explore various plot types
#### Scatter plot of population and gdpPercap
The scatter plot of population and gdpPercap for all countries
```{r}
ggplot(gapminder, aes(pop, gdpPercap)) + 
  geom_point()
```

"filter()" can be used to show the above scatter plot for a single country like Canada
```{r}
gapminder %>% 
  select(country, pop, gdpPercap) %>% 
  filter(country == 'Canada') %>% 
  ggplot(aes(pop, gdpPercap)) + 
  geom_point()
```

Logic "Or" can be used if we want to show results for two or more countries
```{r}
gapminder %>% 
  select(country, pop, gdpPercap) %>% 
  filter(country == 'Canada' | country == "United States") %>% 
  ggplot(aes(pop, gdpPercap)) + 
  geom_point(aes(colour = country))
```

#### Histogram, kernel density plot for gdpPercap
For all countries
```{r}
ggplot(gapminder, aes(gdpPercap)) + 
  geom_histogram(aes(y=..density..), fill = 'blue', bins = 80) + 
  geom_density()
```

Use "filter()" to get the density plot for gdpPercap for Europe
```{r}
gapminder %>% 
  select(continent, gdpPercap) %>% 
  filter(continent == 'Europe') %>%   
  ggplot(aes(gdpPercap)) + 
  geom_histogram(aes(y=..density..), fill = 'blue', bins = 50) + 
  geom_density()
```

#### Box plot for continent and their gdpPercap
For all continents:
```{r}
gapminder %>% 
  select(continent, pop) %>% 
  ggplot(aes(continent, pop)) + 
  scale_y_log10() +
  geom_boxplot(aes(fill = continent, alpha = 0.2)) + 
  geom_jitter(alpha = 0.25)
```

"filter()" can also be used to filter out the continents by intention
For Europe and Americas only:
```{r}
gapminder %>% 
  select(continent, pop) %>% 
  filter(continent == 'Europe' | continent == 'Americas') %>% 
  ggplot(aes(continent, pop)) + 
  scale_y_log10() +
  geom_boxplot(aes(fill = continent, alpha = 0.2)) + 
  geom_jitter(alpha = 0.25)
```

### But I want to do more!
#### Filter command
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

The command is not correct. It truncated half of the result. c("Rwanda", "Afghansitan") makes a list and then the data form is filtered by a list. We want to filter the data by country instead of list. If it is mandatory to use a list, the property is to iterate through the list and filter by every item in the list:

```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

#### Present numerical tables in a more attractive form, such as using knitr::kable().
```{r}
library(knitr)
k = gapminder %>% 
      select(continent, country, year, lifeExp, gdpPercap) %>%
      filter(continent == 'Americas' & year == '2007') %>% 
      kable(align = 'c') 
print(k)
```

#### More dplyr functions for operating on a single table
With "arrange()", the table can be sort by values in the columns.
```{r}
gapminder %>% 
  arrange(lifeExp) %>% 
  filter(continent == 'Africa')
```

use "summarise" to create summaries of Canada population
```{r}
gapminder %>% 
  filter(country == 'Canada') %>% 
  summarise(avg_pop = mean(pop),
            max_pop = max(pop),
            min_pop = min(pop))
```
